---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rpnc250

<!-- badges: start -->
<!-- badges: end -->

The goal of rpnc250 is to provide R functions that produce height, volume, and
biomass estimates using the equations and coefficients from USFS Research Paper
NC-250: Tree volume and biomass equations for the Lake States.

Citation for publication:
Hahn, Jerold T. 1984. Tree volume and biomass equations for the Lake States. Research Paper NC-250. 
St. Paul, MN: U.S. Dept. of Agriculture, Forest Service, North Central Forest Experiment Station

The original publication can be accessed from the
[U.S. Forest Service](https://www.fs.usda.gov/treesearch/pubs/10037).

The equations and species-specific coefficients were copied out of the PDF
version of the publication by Henry Rodman in January 2021. The tables are
[stored in csv format](inst/csv/) in the R package and have been
[translated into package data objects](data-raw/tables.R) for use in the
internal functions. 

The functions for estimating height, volume, and biomass have been
[tested](tests/testthat/test-eq.R) to align with the examples provided in the
original publication.

## Installation

You can install the latest version of rpnc250 from Github with:

``` r
remotes::install_github("SilviaTerra/rpnc250")
```

## Examples

Here are some examples to show a few ways to work with the functions:

First, load a few packages.
```{r libs}
library(rpnc250)
library(magrittr)
```

The package has a set of test trees pulled from some FIA data in northern
Minnesota ([source code](data-raw/test_trees.R)).
```{r test_trees}
head(test_trees)
```

The height equation includes stand basal area as a covariate, so we can
summarize FIA plot-level (four subplots combined) basal area to use for that
purpose. For simplicity, we will assume that site index is constant for all of
the plot locations.

```{r basal_area}
# summarize plot-level basal area
plot_ba <- test_trees %>%
  dplyr::filter(
    !is.na(dbh),
    !is.na(tpa_unadj),
    statuscd == 1
  ) %>%
  dplyr::group_by(
    plt_cn # FIA plot IDs
  ) %>%
  dplyr::summarize(
    bapa = sum(tpa_unadj * 0.005454 * dbh^2),
    .groups = "drop"
  )

# filter the trees and join to plot basal area table
trees_prepped <- test_trees %>%
  dplyr::filter(
    !is.na(dbh),
    statuscd == 1, # only live trees
    !is.na(tpa_unadj)
  ) %>%
  dplyr::left_join(
    plot_ba,
    by = "plt_cn"
  )
```

### Merchantable height
The height equation in the publication generates estimates of height to a given
top diameter (outside bark) provided species, DBH (inches), site index, and
stand basal area.

To estimate height to 4" top we can apply the function `estimate_height`:
```{r height}
# clone prepped tree dataframe
ht_trees <- trees_prepped

ht_trees$height_4 <- estimate_height(
  spcd = ht_trees$spcd,
  dbh = ht_trees$dbh,
  site_index = 65,
  top_dob = 4,
  stand_basal_area = ht_trees$bapa
)
```

It is also possible to use the functions within `dplyr::mutate` calls which
facilitates clean data wrangling workflows.
```{r height_tidy}

ht_trees <- trees_prepped %>%
  dplyr::mutate(
    height_4 = estimate_height(
      spcd = spcd,
      dbh = dbh,
      site_index = 65,
      top_dob = 4,
      stand_basal_area = bapa
    )
  )

```

### Cubic feet
We can summarize total merchantable volume up to 4" top very easily. First, join
the plot-level basal area table to the tree table, then estimate height to 4"
top for each stem. Using those estimates as inputs to the volume equations we
can estimate stem-level volumes.
```{r cuft}
merch_cuft_4 <- trees_prepped %>%
  dplyr::mutate(
    height_4 = estimate_height(
      spcd = spcd,
      dbh = dbh,
      site_index = 65,
      top_dob = 4,
      stand_basal_area = bapa
    ),
    gross_cuft_vol = estimate_volume(
      spcd = spcd,
      dbh = dbh,
      height = height_4,
      vol_type = "cuft"
    )
  )

# plot relationship between diameter and gross volume
merch_cuft_4 %>%
  ggplot2::ggplot(
    ggplot2::aes(
      x = dbh,
      y = gross_cuft_vol,
      color = common_name
    )
  ) +
  ggplot2::geom_point() +
  ggplot2::labs(
    x = "DBH (inches)",
    y = "gross volume to 4\" top (cubic ft)",
    color = "species"
  )
```

### Board feet
To get merchantable volume in board feet (Int'l 1/4) we can set the merchantable
height to 9" for softwood trees, and 10" for hardwood trees. You and I both
know that the large aspen are probably not destined for sawtimber products, but
we can dream.

```{r bdft}
# make table of spcd and hardwood/softwood assignment
spp_groups <- tidyFIA::ref_tables[["species"]] %>%
  dplyr::transmute(
    spcd,
    class = dplyr::case_when(
      major_spgrpcd %in% c(1, 2) ~ "softwood",
      major_spgrpcd %in% c(3, 4) ~ "hardwood"
    )
  )

merch_bdft <- trees_prepped %>%
  dplyr::left_join(
    spp_groups,
    by = "spcd"
  ) %>%
  dplyr::mutate(
    top_dob = dplyr::case_when(
      class == "hardwood" ~ 10,
      class == "softwood" ~ 9
    )
  ) %>%
  dplyr::filter(
    dbh >= top_dob + 1 # make sure dbh is at least as big as top dob
  ) %>%
  dplyr::mutate(
    merch_height = estimate_height(
      spcd = spcd,
      dbh = dbh,
      site_index = 65,
      top_dob = top_dob,
      stand_basal_area = bapa
    ),
    gross_bdft_vol = estimate_volume(
      spcd = spcd,
      dbh = dbh,
      height = merch_height,
      vol_type = "bdft"
    )
  )

merch_bdft %>%
  ggplot2::ggplot(
    ggplot2::aes(
      x = dbh,
      y = gross_bdft_vol,
      color = common_name
    )
  ) +
  ggplot2::geom_point() +
  ggplot2::labs(
    x = "DBH (inches)",
    y = "gross volume (board ft)",
    color = "species"
  )

```

### Biomass
The publication outlines a process for obtaining estimates of green tons of
biomass. This method differs from the methods used by FIA so keep that in mind
if you are attempting to immitate the FIA methodology. The
[FIA methods](https://www.nrs.fs.fed.us/pubs/39555) use the gross cubic foot
volume equations and coefficients from this publication but use a different
process for converting the bole volume estimate into biomass.

```{r biomass}
trees_with_biomass <- trees_prepped %>%
  dplyr::mutate(
    biomass = estimate_biomass(
      spcd = spcd,
      dbh = dbh,
      site_index = 65,
      stand_basal_area = bapa
    )
  )

trees_with_biomass %>%
  ggplot2::ggplot(
    ggplot2::aes(
      x = dbh,
      y = biomass,
      color = common_name
    )
  ) +
  ggplot2::geom_point() +
  ggplot2::labs(
    x = "DBH (inches)",
    y = "biomass (green tons)",
    color = "species"
  )
```

